---
title: "Assignment 4: Data Wrangling"
author: "Aileen Lavelle"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1.  Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Feb 20th @ 5:00pm.

## Set up your session 

#1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.
```{r Load Packages, message = FALSE}
#1a 
library(tidyverse)
library(lubridate)
library(here)
```

# 1b. Check your working directory.

```{r Check working Directory, message = FALSE}
#1b 
getwd()
```

#1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

#2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r, message = FALSE}
#1c 
O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv")
O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv")
PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv")

#2 

#O3_2018
glimpse(O3_2018)
dim(O3_2018)
colnames(O3_2018)
str(O3_2018)

#O3_2019
glimpse(O3_2019)
dim(O3_2019)
colnames(O3_2019)
str(O3_2019)

#PM25_2018
glimpse(PM25_2018)
dim(PM25_2018)
colnames(PM25_2018)
str(PM25_2018)

#PM25_2019
glimpse(PM25_2019)
dim(PM25_2019)
colnames(PM25_2019)
str(PM25_2019)
```


## Wrangle individual datasets to create processed files.

3. Change date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3
#Check class of date column
class(PM25_2019$Date)
class(PM25_2018$Date)
class(O3_2018$Date)
class(O3_2019$Date)

#set as date
PM25_2019$Date <- as.Date(PM25_2019$Date, format = "%m/%d/%y")
PM25_2018$Date <- as.Date(PM25_2018$Date, format = "%m/%d/%y")
O3_2018$Date <- as.Date(O3_2018$Date, format = "%m/%d/%y")
O3_2019$Date <- as.Date(O3_2019$Date, format = "%m/%d/%y")

#Double check that the conversion worked
class(PM25_2019$Date)
class(PM25_2018$Date)
class(O3_2018$Date)
class(O3_2019$Date)

#4
#Selecting Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE columns
#use select to refine data shown
PM25_2019_Refined <- 
  PM25_2019 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

PM25_2018_Refined <- 
  PM25_2018 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

O3_2018_Refined <- 
  O3_2018 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

O3_2019_Refined <- 
  O3_2019 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5
#For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
#use mutate to replace existing columns
PM25_2018_Refined <-
  PM25_2018_Refined %>%
  mutate(AQS_PARAMETER_DESC = "PM2.5")

PM25_2019_Refined <-
  PM25_2019_Refined %>%
  mutate(AQS_PARAMETER_DESC = "PM2.5")

#6
#Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".
#save new datasets with write.csv
write.csv(PM25_2019_Refined, row.names = FALSE, 
          file ="./Data/Processed/EPAair_PM25_NC2019_processed.csv")

write.csv(PM25_2018_Refined, row.names = FALSE, 
          file ="./Data/Processed/EPAair_PM25_NC2018_processed.csv")

write.csv(O3_2018_Refined, row.names = FALSE, 
          file ="./Data/Processed/EPAair_O3_NC2018_processed.csv")

write.csv(O3_2019_Refined, row.names = FALSE, 
          file ="./Data/Processed/EPAair_O3_NC2019_raw.csv")
```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7 
#Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.
#check column names
colnames(PM25_2019_Refined)
colnames(PM25_2018_Refined)
colnames(O3_2018_Refined)
colnames(O3_2019_Refined)
#combine the four datasets using rbind
EPAAirCombined <- rbind(PM25_2019_Refined, PM25_2018_Refined, O3_2019_Refined, O3_2018_Refined)


#8 
#wrangle data
EPAAirModified <- 
  EPAAirCombined %>%
  filter(Site.Name == "Linville Falls"| Site.Name == "Durham Armory"| Site.Name == "Leggett"| Site.Name == "Hattie Avenue"| Site.Name == "Clemmons Middle"| Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co."| Site.Name == "Garinger High School"| Site.Name == "Castle Hayne"| Site.Name == "Pitt Agri. Center"| Site.Name == "Bryson City"| Site.Name =="Millbrook School") %>%
  drop_na(Site.Name) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanlat = mean(SITE_LATITUDE), 
            meanlong = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date)) %>%
  mutate(Year = year(Date))

#9
#Pivot wider to seperate all the variables of AQS_PARAMETER_DESC by mean AQI
EPAAirWider <- 
  EPAAirModified %>%
  pivot_wider(names_from = AQS_PARAMETER_DESC, values_from = meanAQI)

#10
#call the dimesions of the new tidy dataset
dim(EPAAirWider)

#11
write.csv(O3_2019_Refined, row.names = FALSE, 
          file ="./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")
```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r}

#12
#split-apply-combine strategy to generate a summary data frame
EPA.Air.summaries <- 
  EPAAirWider %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(MeanOzone = mean(Ozone), 
            MeanPM2.5 = mean(PM2.5)) %>% 
  drop_na(MeanOzone) 
  
#13
#call for data dimensions
dim(EPA.Air.summaries)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: 
