---
title: "Assignment 3: Data Exploration"
author: "Aileen Lavelle"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Set working directory and load packages,  message = FALSE}
#set working directory
getwd()
setwd('/Users/aileen/Desktop/Duke/Environmental Data Analytics/EDA_Spring_2023_corrected/Data/Raw')

#load necessary packages
library(tidyverse)
library(lubridate)
library(dplyr)
library(RColorBrewer)
library(ggplot2)

#upload datasets and set as factors
read.csv("ECOTOX_Neonicotinoids_Insects_raw.csv")
read.csv("NEON_NIWO_Litter_massdata_2018-08_raw.csv")
Neonics <- read.csv("ECOTOX_Neonicotinoids_Insects_raw.csv")
Litter <- read.csv("NEON_NIWO_Litter_massdata_2018-08_raw.csv")
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are one of the most commonly used insecticides in the U.S. Researchers have attributed to the loss of pollinators in the U.S.to the use of neonicotinoids. We would be interested in the ecotoxicology of neonicotinoids on insects because it can give us a better understanding on the mechanisms behind colony-collapse disorder and other diseases to pollinators. Ultimately, the loss of pollinators means the loss of crops but also the possible loss of natural biodiversity of flowering plants. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying the litter and woody debris that falls to the ground in forests is important because these substrates act as habitat for insects and are key to carbon cycling. Understanding the quantity of habitat for insects might be important for integrating the ecotoxicity data. Additionally, woody debris data may be used to estimate annual Aboveground Net Primary Productivity (ANPP) and aboveground biomass at plot, site, and continental scales.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected from elevated and ground traps
 2. Mass data for each collection event are measured separately for the following functional groups, to an accuracy of 0.01 grams.
 3. Trap placement within plots may be either targeted or randomized, depending on the vegetation.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?
> Answer: 4623 rows, 30 columns

```{r Summary of ECOTOX neonicotinoid dataset}
#Get dimensions of ECOTOX neonicotinoid dataset
##Rows, columns
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Summary of Neonicotinoid Effects}
#creating new function to convert characters to factors
Neonicsnew <-Neonics%>%
        mutate_all(as.factor)
#using summary command on Neonicsnew
summary(Neonicsnew$Effect)
##looking at kinds of population effects 
summary(Neonicsnew$Effect.Measurement)
```
> Answer: Population effects (1803 occurances) and mortality (1493 occurances) were the greatest effects of the neonictinoids. More specially in the population effects, abundance and mortality were the most common effects (1699 and 1294 occurances respectively). It is important to understand how common mortality is as result of Neonicotinoids, because it can indicate the non-selective nature of the insecticide and the severity of the insecticide on pollinators and other insects that provide ecosystem services. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Summary of commonly studied species}
#Summary of common name column of the dataset variables defined as factors 
summary(Neonicsnew$Species.Common.Name)
```

> Answer: The most common species studied were the honey bee (667), the Parasitic Wasp (285), Buff Tailed Bumblebee (183), Carniolan Honey Bee (152), Bumble Bee (140), and the Italian Honeybee (113). Bees and wasps are related as they are both in the order Hymenoptera. All of the insects in this order are winged. Additionally all bees and social wasps pollinate flowers,can both sting and build nests. These species are of special insterests to humans because we rely on pollinators for our crops, flowers and overall ecosystem health. Other insects do different ecosystem services, but a insecticide that results in the loss of pollinators will be detrimental to crop yields. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Checking class of "Conc.1..Author."}
#class function to determine data type of "Conc.1..Author." column
class("Conc.1..Author.")
```

> Answer:The Conc.1..Author.clumn is a character class. This column is not numeric because this is a qualitative variable describing the kind of concentration type of the chemical applied. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Plot number of studies by publication year}
#Creating a plot of number of studies by publication year
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Plot number of studies by publication year and change color}
#checking NAs
is.na(Neonics$Test.Location)
#Removing NAs
NeonicsComplete <- na.omit(Neonics$Test.Location)
#Plot Neonics with Publication.Year and Test.Location in different colors
ggplot(Neonics) + geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location is the lab from 2000-2005 and 2010-2020, inbetween these times in the study, the field natural was the most common from 2005-2010 as well as earlier in the study from the mid 1990s to 2000. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Endpoint Counts}
#Plot Endpoint Counts
ggplot(Neonicsnew, aes(x = Neonicsnew$Endpoint)) +
  geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: The most common end points are NOEL and LOEL. NOEL is no-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test. LOEL is lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Converting collectDate to date}
#check data class
class(Litter$collectDate)
#character is the output
#Convert to date
Litter$collectDate <- as.Date(Litter$collectDate, format = "%y/%m/%d") 
#Check that it is now a date
class(Litter$collectDate)
#Fix earlier dates with create.early.dates function
create.early.dates <- (function(d) {
       paste0(ifelse(d > 191226,"19","20"),d)
       })
#combinig previous function with fixing earlier dates
Litter$collectDate <- create.early.dates(Litter$collectDate)
#Check dates
summary(Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y%m%d"))

#use unique function to look at August 2018
unique(Litter)
#August 6th 2018 & August 2nd 2018
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Unique Vs. Summary}
#Using unique function for Niwot Ridge
unique(Litter)
#188 plots were sampled at Niwot Ridge
#Convert all the characters in Litter to factors
Litternew <-Litter%>%
        mutate_all(as.factor)
#double check outcome from "unique" function
summary(Litternew$siteID)
#188 plots were sampled at Niwot Ridge
```

> Answer: The unique function only returns unique rows in a dataframe while the summary function accumulates all unique values across columns based on values from one column in a short condensed manner. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Bar Graph of Functional Groups}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Boxplot and Violoin plot of dryMass by functionalGroup}
#boxplot of dryMass by functionalGroup
ggplot(Litter, aes(x = dryMass, y = functionalGroup)) +
  geom_boxplot() 
#violin plot of dryMass by functionalGroup
ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup), draw_quantiles = c(0.25, 0.5, 0.75))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplot in this case is better at showing the quantiles of the data and as result the overall spread of the data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles had the highest biomass at these sites and had a very large range of dry mass. Mixed functional groups followed by Twig/branches also had high dry biomass.

```{r page formatting, include=FALSE}
#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
#debugging
options(tinytex.verbose = TRUE)
```
```
