---
title: "Assignment 2: Coding Basics"
author: "Aileen Lavelle"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq(1,100) #use the seq function to see the vector of 1-100

#2. 
#setting "a" as 1-100
a <- seq(1,100)
#calculate mean of "a"
mean(a) 
mean_a <- mean(a)
mean_a
#calculate median of "a"
median(a) 
median_a <- median(a)
median_a

#3. 
#seeing if mean is larger than median
mean_a > median_a

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#name vector of test scores as "student_scores"
student_scores <- c(89, 43, 100, 25)
student_scores

#conditional statement on pass/fail
Fail <- ifelse(student_scores<50, TRUE, FALSE)
Fail

#create vector of student names
student_names <-c("Maeve", "Laura", "Ally", "Aileen")
student_names

#convert vector of student names to a data frame
student_names <-as.data.frame(student_names)
student_names

# add rows using cbind()
df <-cbind(student_names,student_scores,Fail)
df
#Checking to make sure df is now a dataframe
is.data.frame(df)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can contain different data types, ie. characters, numbers, factors and times. A matrix can only contain a single type of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
Fail <- ifelse(student_scores<50, print("TRUE"), print("FALSE"))
Fail

Failed_Students <- function(student_scores) {
  if(x < 50) {
    print("TRUE")
  } 
  else if (x > 50) {
    print("FALSE")
  }
  else {
    x
  }
}
Failed_Students
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The option of if and else both work. The combined ifelse will display all of the command and output. 
